### 2018-02-14 chapters/02-synth.tex
Edit jogo do bicho chapter

The \texttt{R} code below replicates all statistical analyses and graphs included in this chapter. The original data files as well as the final dataset are available at \href{https://github.com/danilofreire/homicides-sp-synth}{https://github.com/danilofreire/homicides-sp-synth}.

>>>>

The \texttt{R} code below replicates all statistical analyses and graphs included in this chapter. The original data files as well as the final dataset are available at \url{https://github.com/danilofreire/homicides-sp-synth}.

### 2018-02-14 chapters/03-bicho.tex
add R code to synth
Edit jogo do bicho chapter

\begin{figure}[!htbp]
	\centering
	\begin{minipage}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth, height=6cm]{images/bicho01.jpg}
	\end{minipage}
	\hfill
	\begin{minipage}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth, height=6cm]{images/bicho02.jpg}
	\end{minipage}
	\caption{Left: cartoon of the Baron of Drummond and the animals of the \emph{jogo do bicho} (1896). Right: entry ticket to Rio's Zoological Garden that allowed the bearer to join the raffle. Sources: Instituto Histórico e Geográfico Brasileiro, Rio de Janeiro, Revista Illustrada, ano 21, no. 718 (1896) and Museu da Imagem e do Som, Rio de Janeiro. Reproduced in \citet[35--36]{chazkel2011laws}.}
	\label{fig:barao}
\end{figure}

Moreover, the animal game plays a crucial role in the expansion of Rio de Janeiro's Carnival Parade, a popular festivity synonymous with Brazil at home and abroad \citep{araujo2003carnaval,costa2001100,da1973carnaval, da1979carnavais,vianna1995misterio}. \emph{Bicheiros} donate hefty sums to `samba schools' to gather support of poor communities and, no less importantly, to co-opt local politicians attracted by the financial and electoral gains offered by the festival \citep{cavalcanti2006carnaval, queiroz1992carnaval}. This patron-client relationship has been proven effective: in 2016, the Carnival generated about USD 900 million in revenue and the Rio de Janeiro state received more than one million tourists.\footnote{Data provided by the Brazilian government. See \url{https://goo.gl/XMcbTM} (in Portuguese). Access: December 2016.}

@@

The remainder of this article proceeds as follows. Section \ref{sec:organisation} presents a brief historical overview of the \emph{jogo do bicho}. It describes the necessary conditions for the emergence of the game and presents its basic organisational structure. Section \ref{sec:governance} details the \emph{jogo do bicho}'s governance mechanisms, particularly the strategies employed by vendors to increase trust in markets that operate at the margins of the law. Section \ref{sec:capture} discusses the links between illegal gambling markets, samba schools and the Brazilian state. Section \ref{sec:conclusion3} offers some concluding remarks.

\section{\emph{Jogo do Bicho} as an Emergent Institution}
\label{sec:organisation}

@@

\begin{figure}[!htbp]
	\begin{tikzpicture}[->,>=stealth',shorten >=4pt,auto,node distance=3cm, semithick][h]
		% nodes %
		\node[circle,fill,inner sep=3pt,label=above right: \shortstack{Urban \\ Poverty}] (p) {};
		\node[circle,fill,inner sep=3pt,label=right: \shortstack{Jogo \\ do Bicho}, right = 8 of p] (y) {};
		\node[circle,fill,inner sep=3pt,label=below left: {Immigration}, below = 3 of p] (i) {};
		\node[circle,fill,inner sep=3pt,label=above:{Money Supply}, above = 3 of p] (m) {};
		\node[circle,fill,inner sep=3pt,label=below right:{Social Ties}, right = 3 of i] (s) {};
		\node[circle,fill,inner sep=3pt,label=above:{Weak Law Enforcement}, above = 3 of y] (l) {};
		\node[circle,fill=white, draw, outer sep=0pt, inner sep=3pt,label= left: \shortstack{Correlated \\ Background Factors}, above left = 2 of p] (u) {};
		% edges %
		\draw[->, line width = 1.2] (p) -- (y);
		\draw[->, line width = 1] (s) -- (y);
		\draw[->, line width = 1] (i) -- (s);
		\draw[->, line width = 1.2] (l) -- (y);
		\draw[->, line width = 1] (m) -- (y);
		\draw[->, line width = 1.2] (i) -- (p);
		\draw[->, dashed, line width = 1] (u) -- (p);
		\draw[->, dashed, line width = 1] (u) -- (m);
	\end{tikzpicture}
	\caption{Directed Acyclic Graph -- Explanatory Variables for the \emph{Jogo do Bicho}}
	\label{fig:dag}
\end{figure}

@@

\begin{figure}[!htbp]
	\centering
	\begin{minipage}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth, height=6cm]{images/bicho03.jpg}
	\end{minipage}
	\hfill
	\begin{minipage}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth, height=6cm]{images/bicho04.jpg}
	\end{minipage}
	\caption{\emph{Jogo do bicho} betting sites in 1917 and in 2011. The picture on the right shows a \emph{bicheiro}, a street-corner vendor. Sources: \citet{alecrim2012bicho} and \citet{ferrarini2011bicho}.}
	\label{fig:banca1917}
\end{figure}

@@

The \emph{jogo do bicho} is a repugnant market. Individuals that like to gamble cannot do so because of strong moral objections from outsiders \citep{brisset2016marche, roth2007repugnance, satz2010some, zelizer1979morals}. As early as in 1890, Brazilian public authorities positioned themselves against the \emph{jogo do bicho} arguing that `{[}\ldots{}{]} this type of amusement is prejudicial to the interests of the unwise, who are naively seduced by the deceptive hope of uncertain lucre' \citep[544]{chazkel2007beyond}. In 1941, the government banned the animal game;\footnote{See: \url{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del3688.htm} (in Portuguese). Access: December 2016.} five years later, it prohibited all games of chance.\footnote{The 1946 decree stated that gambling was `harmful to morality and the good customs', hence `[\dots] the repression against games of chance [was] an imperative of the universal consciousness'. The text can be read at: \url{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del9215.htm} (in Portuguese). Access: December 2016.} The \emph{jogo do bicho}, casinos and bingos remain illegal in the country. Recent estimations show that the prohibition of the \emph{jogo do bicho} have prevented the state from earning BRL 15 to BRL 20 billion (USD 4.5 to USD 6 billion) per year in expected taxation revenues, aside from the subjective utility losses for players. \citep{congressoemfoco2015bicho, fsp2016legalizarbicho}.

@@

\begin{figure}[!htbp]
	\centering
	\includegraphics[width=\textwidth, height=6cm]{images/bicho06.jpg}
	\caption{Results of a \emph{jogo do bicho} draw from 09 January 2016. `Federal' means that the winning numbers were drawn by the federal government lottery. The first prize was group 14, the cat. Source: Unknown. Available at: \url{https://goo.gl/6PHV8u}. Access: December 2016.}
	\label{fig:federal}
\end{figure}

@@

\begin{figure}[!htbp]
	\centering
	\includegraphics[width=\textwidth, height=6cm]{images/bicho05.jpg}
	\caption{\emph{Jogo do bicho} results are fixed on light poles in Rio de Janeiro. Source: \citet{gomes1998bicho}.}
	\label{fig:poste}
\end{figure}

@@

\begin{figure}[!htbp]
	\centering
	\includegraphics[width=.8\textwidth, height=8cm]{images/bicho07.jpg}
	\caption{Castor de Andrade is shown celebrating after the samba school he sponsored, Mocidade Independente, won the Carnival parade in 1996. He was surrounded by colleagues and police officers. Andrade was the founding president of LIESA (1984--1985) and the wealthiest \emph{bicheiro} of Rio de Janeiro at the time. Source: Folha Imagem. Reproduced in \citet[139]{misse2007illegal}.}
	\label{fig:castor}
\end{figure}

@@

\subsection{Political Support}

@@

\begin{figure}[!htbp]
	\centering
	\includegraphics[width=.6\textwidth, height=8cm]{images/bicho08.jpg}
	\caption{Political advertising for Abraãozinho David (right), nephew of the \emph{jogo do bicho} banker Aniz Abraão David (third from left to right). The banner reads: `The Candidates of the Abraão Family: Fighting for the People is a Family Heritage'. Source: \citet{extra2012aniz}.}
	\label{fig:aniz}
\end{figure}

@@

\section{Conclusion}
\label{sec:conclusion3}

>>>>

% \begin{figure}[!htbp]
%  \centering
%  \begin{minipage}[b]{0.45\textwidth}
%   \includegraphics[width=\textwidth, height=6cm]{images/bicho01.jpg}
%  \end{minipage}
%  \hfill
%  \begin{minipage}[b]{0.45\textwidth}
%   \includegraphics[width=\textwidth, height=6cm]{images/bicho02.jpg}
%  \end{minipage}
%  \caption{Left: cartoon of the Baron of Drummond and the animals of the \emph{jogo do bicho} (1896). Right: entry ticket to Rio's Zoological Garden that allowed the bearer to join the raffle. Sources: Instituto Histórico e Geográfico Brasileiro, Rio de Janeiro, Revista Illustrada, ano 21, no. 718 (1896) and Museu da Imagem e do Som, Rio de Janeiro. Reproduced in \citet[35--36]{chazkel2011laws}.}
%  \label{fig:barao}
% \end{figure}

Moreover, the animal game plays a crucial role in the expansion of Rio de Janeiro's Carnival Parade, a popular festivity synonymous with Brazil at home and abroad \citep{araujo2003carnaval,costa2001100,da1973carnaval, da1979carnavais,vianna1995misterio}. \emph{Bicheiros} donate hefty sums to `samba schools' to gather support of poor communities and, no less importantly, to co-opt local politicians attracted by the financial and electoral gains offered by the festival \citep{cavalcanti2006carnaval, queiroz1992carnaval}. This patron-client relationship has been proven effective: in 2016, the Carnival generated about USD 900 million in revenue and the Rio de Janeiro state received more than one million tourists.\footnote{Data provided by the Brazilian government. See \href{https://goo.gl/XMcbTM}{https://goo.gl/XMcbTM} (in Portuguese). Access: December 2016.}

@@

The remainder of this article proceeds as follows. Section \ref{sec:organisation} presents a brief historical overview of the \emph{jogo do bicho}. It describes the necessary conditions for the emergence of the game and presents its basic organisational structure. Section \ref{sec:governance} details the \emph{jogo do bicho}'s governance mechanisms, particularly the strategies employed by vendors to increase trust in markets that operate at the margins of the law. Section \ref{sec:capture} discusses the links between illegal gambling markets, samba schools and the Brazilian state. Section \ref{sec:conclusion} offers some concluding remarks.

\section{\emph{Jogo do Bicho} as an Emergent Institution}
\label{sec:institution}

@@

\begin{figure}[!htbp]
 \begin{tikzpicture}[->,>=stealth',shorten >=4pt,auto,node distance=3cm, semithick][h]
  % nodes %
  \node[circle,fill,inner sep=3pt,label=above right: \shortstack{Urban \\ Poverty}] (p) {};
  \node[circle,fill,inner sep=3pt,label=right: \shortstack{Jogo \\ do Bicho}, right = 8 of p] (y) {};
  \node[circle,fill,inner sep=3pt,label=below left: {Immigration}, below = 3 of p] (i) {};
  \node[circle,fill,inner sep=3pt,label=above:{Money Supply}, above = 3 of p] (m) {};
  \node[circle,fill,inner sep=3pt,label=below right:{Social Ties}, right = 3 of i] (s) {};
  \node[circle,fill,inner sep=3pt,label=above:{Weak Law Enforcement}, above = 3 of y] (l) {};
  \node[circle,fill=white, draw, outer sep=0pt, inner sep=3pt,label= left: \shortstack{Correlated \\ Background Factors}, above left = 2 of p] (u) {};
  % edges %
  \draw[->, line width = 1] (p) -- (y);
  \draw[->, line width = 1] (s) -- (y);
  \draw[->, line width = 1] (i) -- (s);
  \draw[->, line width = 1] (l) -- (y);
  \draw[->, line width = 1] (m) -- (y);
  \draw[->, line width = 1] (i) -- (p);
  \draw[->, dashed, line width = 1] (u) -- (p);
  \draw[->, dashed, line width = 1] (u) -- (m);
 \end{tikzpicture}
 \caption{Directed Acyclic Graph -- Explanatory Variables for the \emph{Jogo do Bicho}}
 \label{fig:dag}
\end{figure}

@@

% \begin{figure}[!htbp]
%  \centering
%  \begin{minipage}[b]{0.45\textwidth}
%   \includegraphics[width=\textwidth, height=6cm]{images/bicho03.jpg}
%  \end{minipage}
%  \hfill
%  \begin{minipage}[b]{0.45\textwidth}
%   \includegraphics[width=\textwidth, height=6cm]{images/bicho04.jpg}
%  \end{minipage}
%  \caption{\emph{Jogo do bicho} betting sites in 1917 and in 2011. The picture on the right shows a \emph{bicheiro}, a street-corner vendor. Sources: \citet{alecrim2012bicho} and \citet{ferrarini2011bicho}.}
%  \label{fig:banca1917}
% \end{figure}

@@

The \emph{jogo do bicho} is a repugnant market. Individuals that like to gamble cannot do so because of strong moral objections from outsiders \citep{brisset2016marche, roth2007repugnance, satz2010some, zelizer1979morals}. As early as in 1890, Brazilian public authorities positioned themselves against the \emph{jogo do bicho} arguing that `{[}\ldots{}{]} this type of amusement is prejudicial to the interests of the unwise, who are naively seduced by the deceptive hope of uncertain lucre' \citep[544]{chazkel2007beyond}. In 1941, the government banned the animal game;\footnote{See: \href{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del3688.htm}{http://www.planalto.gov.br/ccivil\_03/decreto-lei/Del3688.htm} (in Portuguese). Access: December 2016.} five years later, it prohibited all games of chance.\footnote{The 1946 decree stated that gambling was `harmful to morality and the good customs', hence `[\dots] the repression against games of chance [was] an imperative of the universal consciousness'. The text can be read at: \href{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del9215.htm}{http://www.planalto.gov.br/ccivil\_03/decreto-lei/Del9215.htm} (in Portuguese). Access: December 2016.} The \emph{jogo do bicho}, casinos and bingos remain illegal in the country. Recent estimations show that the prohibition of the \emph{jogo do bicho} have prevented the state from earning BRL 15 to BRL 20 billion (USD 4.5 to USD 6 billion) per year in expected taxation revenues, aside from the subjective utility losses for players. \citep{congressoemfoco2015bicho, fsp2016legalizarbicho}.

@@

% \begin{figure}[!htbp]
%  \centering
%  \includegraphics[width=\textwidth, height=6cm]{images/bicho06.jpg}
%  \caption{Results of a \emph{jogo do bicho} draw from 09 January 2016. `Federal' means that the winning numbers were drawn by the federal government lottery. The first prize was group 14, the cat. Source: Unknown. Available at: \href{https://goo.gl/6PHV8u}{https://goo.gl/6PHV8u}. Access: December 2016.}
%  \label{fig:federal}
% \end{figure}

@@

% \begin{figure}[!htbp]
%  \centering
%  \includegraphics[width=\textwidth, height=6cm]{images/bicho05.jpg}
%  \caption{\emph{Jogo do bicho} results are fixed on light poles in Rio de Janeiro. Source: \citet{gomes1998bicho}.}
%  \label{fig:poste}
% \end{figure}

@@

% \begin{figure}[!htbp]
%  \centering
%  \includegraphics[width=.8\textwidth, height=8cm]{images/bicho07.jpg}
%  \caption{Castor de Andrade is shown celebrating after the samba school he sponsored, Mocidade Independente, won the Carnival parade in 1996. He was surrounded by colleagues and police officers. Andrade was the founding president of LIESA (1984--1985) and the wealthiest \emph{bicheiro} of Rio de Janeiro at the time. Source: Folha Imagem. Reproduced in \citet[139]{misse2007illegal}.}
%  \label{fig:castor}
% \end{figure}

@@

\subsection{Political Support}
\label{sub:support}

@@

% \begin{figure}[!htbp]
%  \centering
%  \includegraphics[width=.6\textwidth, height=8cm]{images/bicho08.jpg}
%  \caption{Political advertising for Abraãozinho David (right), nephew of the \emph{jogo do bicho} banker Aniz Abraão David (third from left to right). The banner reads: `The Candidates of the Abraão Family: Fighting for the People is a Family Heritage'. Source: \citet{extra2012aniz}.}
%  \label{fig:aniz}
% \end{figure}

@@

\section{Conclusion}
\label{sec:conclusion}
